The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-28 paper/paper.tex
paper + app folder added, vignette updated

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% need this for pandoc %
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for \widthof, \maxof
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%-------------------------------------------------------------------
% R MARKDOWN
%-------------------------------------------------------------------
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
  \author{
            Dewi Amaliah\thanks{Corresponding author, \href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}}\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Dianne Cook\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Emi Tanaka\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Kate Hyde\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Nicholas Tierney\\
      Telethon Kids Institute, Perth, WA, }
        \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called \texttt{yowie}. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users.
\end{abstract}

\noindent%
{\it Keywords:}  Data cleaning; Data tidying; Longitudinal data; NLSY79; Open data; Initial data analysis; Outlier detection; Robust linear regression
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al. 2020).

@@

Our process of cleaning builds heavily on the tidyverse approach (H. Wickham, Averick, et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data does not comply these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (H. Wickham, Averick, et al. 2019), \texttt{tidyr} (H. Wickham 2020b) to pivot the data into long form, \texttt{dplyr} (H. Wickham, François, et al. 2020), and \texttt{stringr} (H. Wickham 2019) for creating new variables, and levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.

@@

According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

@@

Since we work with longitudinal data, we build the model for each ID instead of the overall data. The robust mixed model could be the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. Thus, the model for each ID is built utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (H. Wickham 2020b) and \texttt{purrr} (Henry and Wickham 2020), respectively.

@@

To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data. We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}.

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. It implies that the fluctuation can still be observed in the data after the treatment. However, the large spikes, which are considered ``erroneous outliers'', are already eliminated from the data. Hence, the model produces a data set with a more reasonable degree of fluctuation.

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/summer_project/summer-wages-refresh/paper/figures/flowchart}

@@

There are two aspects of the original data that make a direct comparison difficult. The time variable provided is ``experience in the workforce''. It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

@@

The entire analysis is conducted using \texttt{R} (R Core Team 2020) in \texttt{rstudio} using these packages: \texttt{tidyverse} (H. Wickham, Averick, et al. 2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (H. Wickham, François, et al. 2020), \texttt{readr} (H. Wickham and Hester 2020), \texttt{tidyr} (H. Wickham 2020b), \texttt{stringr} (H. Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{broom} (Robinson, Hayes, and Couch 2021), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), \texttt{DiagrammeR} (Iannone 2020), \texttt{rsvg} (Ooms 2020), \texttt{webshot} (Chang 2019), \texttt{mgcv} (Wood 2003), \texttt{tsibble} (Wang, Cook, and Hyndman 2020), and \texttt{modelr} (H. Wickham 2020a). The paper are generated using \texttt{knitr} (Xie 2014), \texttt{rmarkdown} (Xie, Dervieux, and Riederer 2020), and \texttt{rticles} (Allaire et al. 2021).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-rticles}{}%
Allaire, JJ, Yihui Xie, R Foundation, Hadley Wickham, Journal of Statistical Software, Ramnath Vaidyanathan, Association for Computing Machinery, et al. 2021. \emph{Rticles: Article Formats for R Markdown}. \url{https://CRAN.R-project.org/package=rticles}.

@@

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{Shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\hypertarget{ref-eliznlsy}{}%
Cooksey, Elizabeth C. 2017. ``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. \url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.

@@

\leavevmode\hypertarget{ref-racismnotrace}{}%
Fullilove, M. T. 1998. ``Comment: Abandoning "Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.

@@

\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. ``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa. 2020. ``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``Robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render Svg Images into Pdf, Png, Postscript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.

@@

\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. ``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

@@

\leavevmode\hypertarget{ref-nlsy79}{}%
The U.S. Bureau of Labor Statistics. n.d. ``National Longitudinal Survey of Youth 1979.'' Available at \url{https://www.nlsinfo.org/content/cohorts/nlsy79} (2021/25/02).

\leavevmode\hypertarget{ref-brolgar}{}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{Brolgar: BRowse over Longitudinal Data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.

@@

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. ``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\hypertarget{ref-plyr}{}%
Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.

@@

\leavevmode\hypertarget{ref-tidyverse}{}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. ``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

@@

\leavevmode\hypertarget{ref-mgcv}{}%
Wood, S. N. 2003. ``Thin-Plate Regression Splines.'' \emph{Journal of the Royal Statistical Society (B)} 65 (1): 95--114.

\leavevmode\hypertarget{ref-knitr}{}%
Xie, Yihui. 2014. ``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

@@

\leavevmode\hypertarget{ref-kableExtra}{}%
Zhu, Hao. 2019. \emph{KableExtra: Construct Complex Table with 'Kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.

\bibliographystyle{Chicago}

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{graphicx}

\title{The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}

\author{
    Dewi Amaliah
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dama0007@student.monash.edu}{\nolinkurl{dama0007@student.monash.edu}}} \\
   \And
    Dianne Cook
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
   \And
    Emi Tanaka
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
   \And
    Kate Hyde
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{} \\
   \And
    Nicholas Tierney
   \\
     \\
   \\
  \texttt{} \\
  }

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called \texttt{yowie}. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users.
\end{abstract}

\keywords{
    Data cleaning; Data tidying; Longitudinal data; NLSY79; Open data; Initial data analysis; Outlier detection; Robust linear regression
  }

@@

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (M. Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M. Huebner et al. 2020).

@@

Our process of cleaning builds heavily on the tidyverse approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data does not comply these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019), \texttt{tidyr} (Wickham 2020b) to pivot the data into long form, \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.

@@

According to Huebner Marianne, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

@@

Since we work with longitudinal data, we build the model for each ID instead of the overall data. The robust mixed model could be the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. Thus, the model for each ID is built utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020b) and \texttt{purrr} (Henry and Wickham 2020), respectively.

@@

To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier,'' we simulate some thresholds and study how they affect the data. We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}.

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. It implies that the fluctuation can still be observed in the data after the treatment. However, the large spikes, which are considered ``erroneous outliers,'' are already eliminated from the data. Hence, the model produces a data set with a more reasonable degree of fluctuation.

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/etan0038/Dropbox/repo/summer-wages-refresh/paper/figures/flowchart}

@@

There are two aspects of the original data that make a direct comparison difficult. The time variable provided is ``experience in the workforce.'' It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

@@

The entire analysis is conducted using \texttt{R} (R Core Team 2020) in \texttt{rstudio} using these packages: \texttt{tidyverse} (Wickham et al. 2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham et al. 2020), \texttt{readr} (Wickham and Hester 2020), \texttt{tidyr} (Wickham 2020b), \texttt{stringr} (Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{broom} (Robinson, Hayes, and Couch 2021), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), \texttt{DiagrammeR} (Iannone 2020), \texttt{rsvg} (Ooms 2020), \texttt{webshot} (Chang 2019), \texttt{mgcv} (Wood 2003), \texttt{tsibble} (Wang, Cook, and Hyndman 2020), and \texttt{modelr} (Wickham 2020a). The paper are generated using \texttt{knitr} (Xie 2014), \texttt{rmarkdown} (Xie, Dervieux, and Riederer 2020), and \texttt{rticles} (Allaire et al. 2021).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-rticles}{}%
Allaire, JJ, Yihui Xie, R Foundation, Hadley Wickham, Journal of Statistical Software, Ramnath Vaidyanathan, Association for Computing Machinery, et al. 2021. \emph{Rticles: Article Formats for r Markdown}. \url{https://CRAN.R-project.org/package=rticles}.

@@

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{Shiny: Web Application Framework for r}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\hypertarget{ref-eliznlsy}{}%
Cooksey, Elizabeth C. 2017. {``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.

@@

\leavevmode\hypertarget{ref-racismnotrace}{}%
Fullilove, M. T. 1998. {``Comment: Abandoning "Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.

@@

\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. {``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa. 2020. {``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.''} \emph{BMC Medical Research Methodology} 20 (1): 61--61.

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. {``Robustlmm: An r Package for Robust Estimation of Linear Mixed-Effects Models.''} \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render SVG Images into PDF, PNG, PostScript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.

@@

\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. {``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

@@

\leavevmode\hypertarget{ref-nlsy79}{}%
The U.S. Bureau of Labor Statistics. n.d. {``National Longitudinal Survey of Youth 1979.''} Available at \url{https://www.nlsinfo.org/content/cohorts/nlsy79} (2021/25/02).

\leavevmode\hypertarget{ref-brolgar}{}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{Brolgar: BRowse over Longitudinal Data Graphically and Analytically in r}. \url{https://github.com/njtierney/brolgar}.

@@

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. {``Robust Regression \textbar{} r Data Analysis Examples.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. {``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\hypertarget{ref-plyr}{}%
Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.

@@

\leavevmode\hypertarget{ref-tidyverse}{}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. {``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

@@

\leavevmode\hypertarget{ref-mgcv}{}%
Wood, S. N. 2003. {``Thin-Plate Regression Splines.''} \emph{Journal of the Royal Statistical Society (B)} 65 (1): 95--114.

\leavevmode\hypertarget{ref-knitr}{}%
Xie, Yihui. 2014. {``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

@@

\leavevmode\hypertarget{ref-kableExtra}{}%
Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'Kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.

\end{CSLReferences}

\bibliographystyle{unsrt}
\bibliography{references.bib}
